\section{Link with "Old" parametricity (2)}
We define a (unary) relational interpretation of types $⟦A⟧$. This
interpretation coincides with the usual relational interpretation of types.
We show that, if $⟦·⟧$ and $∋_i$ are equivalent for free type variables,
then they are equivalent for every type in normal form.

\begin{definition}
We define $⟦A⟧$ by structural induction on $A$.
  \begin{align*}
    ⟦A → B⟧ f & = (x:\proj i A) → ⟦A⟧ x → ⟦B⟧ (f x)\\
    ⟦U⟧ A & = A → U\\
    ⟦A \op j a ⟧ b &= ⟦A⟧ \CP j {\proj i a} b \op j a \param i \\
    ⟦\ip i A P⟧ x & = P x\\
    ⟦x⟧ a & = x \op i a
  \end{align*}
\end{definition}

\providecommand\TO{\overrightarrow}
\providecommand\FROM{\overleftarrow}

\begin{theorem}
For every type $A$, $A \op i x$ is provable iff. $⟦A⟧ x$ is provable.
\end{theorem}
Let us call $\TO A$ the conversion from $A \op i x$ to $⟦A⟧ x$; and $\FROM A$ the other direction.
\begin{proof}
  By structural induction on $A$. Key cases:
  \begin{itemize}
  \item Pi.
    Assume:
    \begin{itemize}
    \item $f : A → B$
    \item $p : (A → B) \op i f$
    \end{itemize}
    We prove $q : (x:\proj i A) → ⟦A⟧ x → ⟦B⟧ (f x)$ with $q x x' = \TO {B[\CP i x {\FROM A x'}]} ((\CP i f
    p \CP i x {\FROM A x'}) \param i)$.

    Right to left: $p = (\fp i f {λx. λx'. \FROM {B[\CP i x {x'}]} (q x (\TO A x'))}) \param i $

    (No need to carry around an environment; the proof given by the
    user ($x'$) is substituted in terms --- the substitutions are
    well-typed, given the {\sc In-Abs} rule.)

  \item Universe.

    Left to right. Assume:
    \begin{itemize}
    \item $A : U$
    \item $P : U \op i A $
    \end{itemize}
    We prove $Q : A → U$ with $Q x = \CP i A P \op i x$.
 

    Right to left: $P = (\ip i A Q) \param i $

  \item Sigma.
    By def.
  \item Param.

    Left to right. Assume:
    \begin{itemize}
    \item $a : A$
    \item $b : A\op j a$
    \item $p : A \op j a \op i b $
    \end{itemize}
    We prove $q' : A \op i \CP j {\proj j a} b \op j a \param i  $ with $\CP i a {\CP j b p} \param j \param i$.
    The result is given by $\TO A q'$.

    Right to left: same principle.
  \end{itemize}
  \item Variable.
    By def.
\end{proof}
