\documentclass[10pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amssymb,amstext,amsmath}
\usepackage{mathpartir}
\usepackage{mytheorems}
\usepackage[a4paper,margin=1.8cm]{geometry}
\usepackage{fancyref}

\newcommand\CC[4]{(#2,_{#1} #3)_{#4}}
\newcommand\CSig[1]{\Sigma^{#1}}
\newcommand\CTimes[3]{#2 ×_{#1} #3}
\newcommand\SW[2]{\mathsf{SW}^{#1}_{#2}}
\newcommand\sw[2]{\mathsf{sw}^{#1}_{#2}}
\newcommand\dom{\mathsf{dom}}
\newcommand\param[1]{\!\cdot\!#1}
\newcommand\pvar[2]{{#1}^{(#2)}}


\input{unicodedefs}

\title{A parametric Type theory}

\author{}
\date{\today}

\begin{document}
\maketitle


\begin{definition}[Syntax of terms and contexts]
  
  \begin{align*}
    t,u,A,B & ::= x ~|~ U ~|~ λx:A. t      ~|~ t u ~|~ (x:A) → B \\
            & ~|~ \CC i t u A  ~|~ t \param i ~|~ \sw {ij} T \\
    \Gamma,\Delta & ::= () ~|~ \Gamma,x:A ~|~ \Gamma,I 
  \end{align*}
\end{definition}
$\CSig i A P$ and $(x:A) ×_i P x$ are other notations for $(A ,_i P)_U$.

 We use $I,J,…$ for finite {\em sets} of colours (order does not matter).

\begin{definition}[Typing relation]
Contexts:
  \begin{mathpar}
    \inferrule[Empty]{~}{() ⊢ }

    \inferrule[Cons]{Γ⊢ \\ Γ ⊢ A }{ Γ,x:A ⊢ }

    \inferrule[CCons]{Γ⊢ \\ Γ ⊢ }{ Γ,I ⊢ }

    \inferrule[NewVar]{Γ⊢ \\ Γ ⊢ A }{ Γ,x:A ⊢ }

    \inferrule[NewColor]{Γ⊢}{ Γ,I ⊢ }

  \end{mathpar}
 
Types:
 \begin{mathpar}
    \inferrule[Universe]{Γ ⊢}{Γ ⊢ U : U}

    \inferrule[Pi]{Γ ⊢ A : U \\ Γ,x:A ⊢ B : U}{Γ ⊢ (x:A) → B : U}

  \end{mathpar}

Terms:
  \begin{mathpar}
    \inferrule[Conv]{Γ ⊢ t:A \\ A = B}{Γ ⊢ t : B}

    \inferrule[Var]{Γ ⊢ A}{Γ,x:A ⊢ x : A}

    \inferrule[Wk]{Γ ⊢ A \\Γ ⊢ t : T}{Γ,x:A ⊢ t : T}

    \inferrule[CWk]{Γ ⊢ t : T}{Γ,I ⊢ t : T}

    \\\inferrule[Lam]{Γ,x:A ⊢ B}{Γ ⊢ λx:A.b : (x:A) → B}
 
    \inferrule[App]{Γ ⊢ t : (x:A) → B[x] \\ Γ ⊢ u : A}{Γ ⊢ t u: B[u] }

    \\\inferrule[Coloring]{Γ,I ⊢ a : T i0 \\Γ,I ⊢ p : T \param i a}{Γ,I,i ⊢ (a,_i p)_T : T}

    \inferrule[Param]{Γ,I,i ⊢ a : T}{Γ,I ⊢ a \param i : T \param i a (i0)}
 
    \inferrule[Swap]{Γ,I,i,j ⊢ T }{Γ,I ⊢ \sw {ij} T : \SW {ij} T}

  \end{mathpar}
\end{definition}
%    \inferrule[Swap]{\dom (σ) = J \\ I ∩ J = ∅ \\ Γ ⊢_{I} T }{Γ ⊢_{I} \sw {σ} T : (x : T)_J → T ∋_{(σ J)} x  }


where $\SW {ij} T = (x:T(i0)(j0)) → (x_j:T(i0) \param j x) → (x_i : T(j0) \param i x) → T \param i \param j x x_j x_i → T \param j \param i x x_i x_j$.

\begin{definition}[Reduction]~

Functions:
\begin{align*}
  (λx:A. u[x]) t &= u[t]  \\
  (f ,_i g)_{(x:A)→ B[x]} a & = (f a(i0) ,_i g a(i0) a \param i)_{B[a]}
\end{align*}

Param:
\begin{align*}
  U\param i T &= T → U \\
  ((x:A) → B[x])\param i f &= (x:A) → (x' : A\param i x) → (B[(x,_i x')_A])\param i (f x) \\
  (λx:A. b[x])\param i &= λx:A(i0). λx':A\param i x. b[\CC i x {x'} A]\param i \\
  (f a)\param i &= f\param i (a (i0)) (a\param i) \\
  (a,_i p)_T\param i  &= p \\
  (a,_i p)_T \param j  &= (a \param j ,_i \sw {i j} T   a(j0)   a \param j   p(j0)   p \param j)_{T \param j ((a ,_i p) j0)}  & (*)
\end{align*}

Swap:
\begin{align*}
  \sw {i j} {U} A A_j A_i A_{ij} a a_i a_j & = A_{ij} a a_j a_i  \\
  \sw {i j} {(x:A) → B[x]} f f_j f_i f_{ij} a a_i a_j a_{ji} & = \sw {i j} {B[((a,_ia_i),_j(a_j,_ia_{ji}))]} (f a)
(f_j a a_j) (f_i a a_i) (f_{ij} a a_j a_i (\sw {j i} A a a_i a_j a_{ji})) \\
  \sw {i j} {(A ,_i P)_U} x x_j x_i x_{ij} & = x_{ij} \\
  \sw {i j} {(A ,_j P)_U} x x_j x_i x_{ij} & = x_{ij} \\
  \sw {i j} {(A ,_k P)_U} (x ,_k x_k) (x_j ,_k x_{jk}) (x_i ,_k x_{ik}) (x_{ij},_k x_{ijk}) & = 
    (\sw {i j} A x  x_j  x_i  x_{ij},_k \sw {i j} {P((x,_jx_j),_i(x_i,_j x_{ij}))} x_k  x_{jk}  x_{ik}  x_{ijk})_X & (**) \\
\end{align*}

%  The last rule might also be written for $T = (A,_k P)_U$  (CONJECTURE)
% $$
% \sw {ij} T = (\sw {ij} A,_k \sw {ij} {(x:A)→ Px})_{\SW {ij} T}
% $$
% Nope: the second part of the pair needs four functions to swap; but the first part is a single function (which also has not quite the right shape).

\begin{itemize}
\item 
(*): $p : T\param i a$ so $p \param j : (T\param i a) \param j p(j0)$. But,
the second component of the pair should be of type $(T \param j ((a ,_i p)
j0))\param i a \param j$.  That is, we need $\sw{ i j} T$ to be a conversion from
$T\param i \param j a(j0) a \param j p(j0)$ to $T \param j\param i a(j0) p(j0) a \param j$.

\item (**): The source type of the swap is
$$W = (x_{ij}:A\param i \param j x x_j x_i) \times_k P\param i \param j x x_j x_i x_{ij} x_k x_{jk}
x_{ik}$$ By computation (using other cases of the definition), all the
occurences of swap disappear in $W$.

The target type is $X$ and $X = (x_{ji}:A \param j\param i x x_i x_j) \times_k
P \param j \param i x x_i x_j x_{ji} x_k x_{ik} x_{jk}$.  One might think that swaps
should be applied to certain variables in $X$, but it is not
the case.  This can be checked by verifying that the following types
for variables match both the source and the target.

\end{itemize}
\end{definition}




\begin{definition}[Conversion]
  \begin{mathpar}
    \inferrule{t x = u} {t = λ x:A.u}

    \inferrule{t (i0) = a \\ t.i = p} {t = \CC i a p T}
  \end{mathpar}
+ congruences.
\end{definition}

 We define substitution between contexts

  \begin{align*}
    σ, δ & ::= ()  |  (σ,x=t)  |  (σ,f)
  \end{align*}

  \begin{mathpar}
    \inferrule {σ : Δ → Γ\\ Δ ⊢ t:Aσ} {(σ,x=t) : Δ → Γ,x:A} 

    \inferrule {σ : Δ → Γ} {(σ,f) : Δ → Γ,J}
  \end{mathpar}
where $f$ is a function of domain $J$ and taking as values $0,1$ or
colours declared in $Δ$ (and such that $j=k$ if $f(j) = f(k)$ is a colour).

The identity substitution $1_{Γ}$ is defined as usual and $(i0)$ is
an abbreviation for $(1_{Γ},i=0)$. Also $b[u]$ is an abbreviation for
$b(1_{Γ},x=u)$.
We define $tσ$ by induction on $t$

\begin{align*}
  Uσ &= U \\
  x(σ,f) &= xσ \\
  x(σ,y=u) &= xσ \\
  x(σ,x=u) &= u \\
  ((x:A)→ B)σ &= (y:Aσ) → B(σ,x=y) \\
  (λ x:A. b)σ &= λ y:Aσ. b(σ,x=y) \\
  (t.i)σ &= (t(σ,i=j)).j \\
  (f a)σ &= fσ  (aσ) \\
  (a,_i b)_T(σ,x=u)  &= (a,_i b)_Tσ  \\
  (a,_i b)_T(σ,f)  &= (a,_i p)_Tσ & i \# f \\
  (a,_i b)_T(σ,f)  &= a(σ,f-i) & f(i) = 0 \\
  (a,_i b)_T(σ,f)  &= (a(σ,f-i),_j b(σ,f-i))_{T(σ,f)} & f(i) = j
\end{align*}

\begin{lemma}\label{lem:term-swap}
  for any $a:T$, $a.j.i = \sw {ij} T a(i0)(j0)~a(i0).j~a(j0).i~a.i.j$.
\end{lemma}
\begin{proof}
  \begin{align*}
    a &= a & \text{refl} \\
    a &= \CC i {a(i0)} {a.i} T & \text{conversion rule} \\
    a.j & = \CC i {a(i0)} {a.i} T .j & \text{congruence} \\
    a.j & = \CC i {a(i0).j} {\sw{ij} T a(i0)(j0)~a(i0).j~a(j0).i~a.i.j} T & \text{reduction} \\
    a.j.i & = \CC i {a(i0).j} {\sw{ij} T a(i0)(j0)~a(i0).j~a(j0).i~a.i.j} T.i & \text{congruence} \\
    a.j.i & = \sw{ij} T a(i0)(j0)~a(i0).j~a(j0).i~a.i.j & \text{subst} \\
\end{align*}
\end{proof}

\subsection{Swap and symmetric groups.}

The function $\sw {ij} T$ has four arguments, but the one which
matters is the last one.  Indeed, the purpose of the term $\sw {ij} T$
is to convert from $T.i.j$ to $T.j.i$, that is, to permute the role of
$i$ and $j$. In the rest of the section we ignore the first three
arguments of $\sw {ij} T$.

\begin{theorem}
For any permutation $σ$, conversion from $T.i_0 … i_n$ to $T.σ(i_0) …
σ(i_n)$ is definable.

\end{theorem}

\begin{proof}
As examples, consider the following cases:
\begin{itemize}
\item $\sw{ij} {T.k~x}$ converts from $T.k.i.j$ to $T.k.j.i$.
\item $\sw{ij} {T}.k$ converts from $T.i.j.k$ to $T.j.i.k$.
\end{itemize}

Let us see now the general case.
Let $s = \sw{i_m~i_{m+1}} {T.i_0⋯T_{m-1} x_1 … x_{2^{m-1}}}.i_{m+2}⋯i_n$.

The effect of $s$ is to convert from $T.i_0 ⋯ i_n$ to
$$T.i_0⋯i_{m-1}.i_{m+1}.i_m .i_{m+2}⋯i_n$$

That is, $s$ performs the permutation of colors $(i_m~i_{m+1})$ over
the domain $i_0 … i_n$. The permutations generated by terms in the
above form are generators of the symmetric group over $i_0 … i_n$.
\end{proof}

In the following, we write $\sw {σ} T$ the swapping term constructed
above for the permutation $σ$. The function $\sw {σ} T$ has $2^{|σ|}$
arguments, of which only the last one matters.

\begin{lemma}
  $\sw{ij} T x x_j x_i (\sw {ji} T x x_j x_i x_{ji}) = x_{ji}$
\end{lemma}
\begin{proof}
  Let $a = \CC j {\CC i x {x_i} {T(j0)}} {\CC i {x_j} {x_{ji}} {T.i}} T$.
  By Lem. \ref{lem:term-swap}, $a.j.i = \sw {ji} T x x_j x_i
  a.j.i$. By performing reductions, we obtain our result.
\end{proof}


\begin{lemma}[$σ_i^2 = 1$]
  We obtain this from first moving the $.k$'s out; then the above lemma, then pushing them back in.
\end{lemma}
\begin{lemma}[$σ_i ∘ σ_j = σ_j ∘ σ_i$ if $abs(j-i) > 1$]
  
\end{lemma}
\begin{proof}
  A consequence of using variable names instead of positions.
\end{proof}
\begin{lemma}
  [$\sigma_i ∘\sigma_{i+1} ∘\sigma_i = \sigma_{i+1} ∘\sigma_i ∘\sigma_{i+1}$]
  $\sw{kj}T.i (\sw{ik}{T.j} (\sw{ij}{T}.k~ x_{ijk})) = \sw{ij}{T.k} (\sw{ki}T.j  (\sw{jk}{T.i} x_{ijk}))$
\end{lemma}
\begin{proof}
  Compute the lhs:
  \begin{align*}
     \sw{kj}T.i (\sw{ik}{T.j} (\sw{ij}{T}.k (a.i.j.k)) ) 
    &= \sw{kj}T.i (\sw{ik}{T.j} (\sw{ij}{T} a.i.j).k ) \\
    &= \sw{kj}T.i (\sw{ik}{T.j} (a.j.i).k ) \\
    &= \sw{kj}T.i (\sw{ik}{T.j} a.j.i.k ) \\
    &= \sw{kj}T.i  (a.j.k.i) \\
    &= (\sw{kj}T  a.j.k).i \\
    &= (a.k.j).i \\
    &= a.k.j.i 
  \end{align*}
  Computing the rhs yields the same expression.
  Let $a$ be a cube constructed such that $a.k.j.i = x_{ijk}$.
\end{proof}
\begin{theorem}
  The generators satisfy the laws of symmetric groups.
\end{theorem}
\begin{proof}
  Combination of the above lemmas.
\end{proof}

\begin{lemma}\label{lem:param-var}
  $x.i.j = x.j.i$
\end{lemma}
\begin{proof}
  By $α$-equivalence of terms.
\end{proof}

Applying parametricity to many variables is useful and (by the above)
does not depend on the names of the variables. Hence we use the
notation $\pvar x k$ for $k$ applications of parametricity to the
variable $x$.

\begin{lemma}\label{lem:sym-two}
  For any variable $x$, $\sw {ij} T x.i.j = x.i.j$.
\end{lemma}
\begin{proof}
  Because $.i$ is a binder, we have $x.i.j = x.j.i$. Then combine this
  result with lemma \ref{lem:term-swap}.
\end{proof}
\begin{corollary}
  $\pvar x k$ generates terms which are invariant under all permutations
  of $1…k$. That is, 

   $$   \text{if~} dom π ⊆ \{1…k\} \text{~then~} \sw {π} {} \pvar x k = \pvar x k$$
\end{corollary}

\begin{lemma}[$U$ is variant under swap] \label{lem:sw-U}
  $\sw {ij} U U.i.j ‌≠ U.j.i$
\end{lemma}
\begin{proof}
By definition.
\end{proof}

\begin{theorem}[inconsistency]
  
\end{theorem}
\begin{proof}
  By lemma \ref{lem:sym-two}, we have
  $$\sw {ij} T x.i.j = x.i.j$$
  Substitute $U$ for $x$:
  $$\sw {ij} U U.i.j = U.i.j$$
  Which is not consistent with lemma \ref{lem:sw-U}.
\end{proof}

\subsection{Normal forms}

\begin{align*}
  n,m & ::= \pvar x k ~|~ n t \\
  t,u,A,B & ::= U ~|~ λx:A. t      ~|~ \sw {π} {} n ~|~  (x:A) → B 
             ~|~ \CC i t u A
\end{align*}


\subsection{Semantics}

- Nominality; dependence on a finite number of colors.


\end{document}
