\documentclass[10pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amssymb,amstext,amsmath}
\usepackage{mathpartir}
\usepackage{mytheorems}
\usepackage[a4paper,margin=1.8cm]{geometry}
\usepackage{fancyref}

\newcommand\CC[4]{(#2,_{#1} #3)_{#4}}
\newcommand\CSig[1]{\Sigma^{#1}}
\newcommand\CTimes[3]{#2 ×_{#1} #3}
\newcommand\SW[2]{\mathsf{SW}^{#1}_{#2}}
\newcommand\sw[2]{\mathsf{sw}^{#1}_{#2}}
\newcommand\dom{\mathsf{dom}}
\newcommand\param[1]{\!\cdot\!#1}
\newcommand\pvar[2]{{#1}^{(#2)}}


\input{unicodedefs}

\title{A parametric Type theory}

\author{}
\date{\today}

\begin{document}
\maketitle


\begin{definition}[Syntax of terms and contexts]
  
  \begin{align*}
    t,u,A,B & ::= x ~|~ U ~|~ λx:A. t      ~|~ t u ~|~ (x:A) → B \\
            & ~|~ \CC i t u A  ~|~ t \param i  \\
    \Gamma,\Delta & ::= () ~|~ \Gamma,x:A ~|~ \Gamma,I 
  \end{align*}
\end{definition}
$\CSig i A P$ and $(x:A) ×_i P x$ are other notations for $(A ,_i P)_U$.

We use $I,J,…$ for finite {\em sets} of colours (order does not matter).

\begin{definition}[Typing relation]
Contexts:
  \begin{mathpar}
    \inferrule[Empty]{~}{() ⊢ }

    \inferrule[NewVar]{Γ⊢ \\ Γ ⊢ A }{ Γ,x:A ⊢ }

    \inferrule[NewColor]{Γ⊢}{ Γ,I ⊢ }

  \end{mathpar}
 
Types:
 \begin{mathpar}
    \inferrule[Universe]{Γ ⊢}{Γ ⊢ U : U}

    \inferrule[Pi]{Γ ⊢ A : U \\ Γ,x:A ⊢ B : U}{Γ ⊢ (x:A) → B : U}
  \end{mathpar}

Terms:
  \begin{mathpar}
    \inferrule[Conv]{Γ ⊢ t:A \\ A = B}{Γ ⊢ t : B}

    \inferrule[Var]{Γ ⊢ A}{Γ,x:A ⊢ x : A}

    \inferrule[Wk]{Γ ⊢ A \\Γ ⊢ t : T}{Γ,x:A ⊢ t : T}

    \inferrule[CWk]{Γ ⊢ t : T}{Γ,I ⊢ t : T}

    \\\inferrule[Lam]{Γ,x:A ⊢ B}{Γ ⊢ λx:A.b : (x:A) → B}
 
    \inferrule[App]{Γ ⊢ t : (x:A) → B[x] \\ Γ ⊢ u : A}{Γ ⊢ t u: B[u] }

    \\\inferrule[Color-intro]{Γ,I ⊢ a : T i0 \\Γ,I ⊢ p : T \param i a}{Γ,I,i ⊢ (a,_i p)_T : T}

    \inferrule[Color-elim]{Γ,I,i ⊢ a : T}{Γ,I ⊢ a \param i : T \param i a (i0)}

  \end{mathpar}
\end{definition}

\begin{definition}[Reduction]~

Functions:
\begin{align*}
  (λx:A. u[x]) t &= u[t]  \\
  (f ,_i g)_{(x:A)→ B[x]} a & = (f a(i0) ,_i g a(i0) a \param i)_{B[a]}
\end{align*}

Param:
\begin{align*}
  U\param i T &= T → U \\
  ((x:A) → B[x])\param i f &= (x:A) → (x' : A\param i x) → (B[(x,_i x')_A])\param i (f x) \\
  \CSig i A P \param i x  &= P x\\
  \CSig i A P \param j (x,i {x_i}) &= (x_j : A \param j x) ×_i (P \param j x x_j x_i) \\
  (λx:A. b[x])\param i &= λx:A(i0). λx':A\param i x. b[\CC i x {x'} A]\param i \\
  (f a)\param i &= f\param i (a (i0)) (a\param i) \\
  (a,_i p)_T\param i  &= p \\
  (a,_i p)_{\CSig i A P} \param j  &= (a \param j ,_i p \param j) \\
  (a,_i p)_{\CSig j A P} \param j  &= (a \param j ,_i p \param j) & ???\\
\end{align*}
\end{definition}

\begin{definition}
  A type (term) $T$ is called $i$-stable when, for any $j$: $T\param i \param j x y z = T\param j \param i x z y $.

\end{definition}
\begin{conjecture}
  Only $\CSig i A P$ is  $i$-stable.
\end{conjecture}
In particular, $U$ is not $i$-stable.

\begin{conjecture}[Conversion]
  \begin{mathpar}
    \inferrule{t x = u} {t = λ x:A.u}

    \inferrule{i ≠ j \\ A(j0) = B(i0) \\ A.j = Q(i0) \\ B.i = P(j0) \\ flip (P.j) = Q.i} {\CSig i A P = \CSig j B Q}

    % \inferrule{t (i0) = a \\ t.i = p} {t = \CC i a p {\CSig i A P}}

    \inferrule{t (i0) = u (i0) \\ t.i = u.i \\ \text{If $T$ is $i$-stable}} {t = u : T}

    \inferrule{???} {t.i = u}
  \end{mathpar}
+ congruences.

One \emph{cannot} deduce $\CSig i T(i0) T.i = T$. Hence in particular the last rule is valid only in the case where the type of $t$ is either $\CSig i A P$ or $\CSig j B Q$.
In the second case, the rule subsumes the following particular case:

    \inferrule{a(j0) = b(i0) \\ a.j = q(i0) \\ b.i = p(j0) \\ p.j = q.i} {\CC i a p {\CSig i A P} = \CC j b q {\CSig i B Q}}
\end{conjecture}

\begin{definition}
 We define substitution between contexts

  \begin{align*}
    σ, δ & ::= ()  |  (σ,x=t)  |  (σ,f)
  \end{align*}

  \begin{mathpar}
    \inferrule {σ : Δ → Γ\\ Δ ⊢ t:Aσ} {(σ,x=t) : Δ → Γ,x:A} 

    \inferrule {σ : Δ → Γ} {(σ,f) : Δ → Γ,J}
  \end{mathpar}
where $f$ is a function of domain $J$ and taking as values $0,1$ or
colours declared in $Δ$ (and such that $j=k$ if $f(j) = f(k)$ is a colour).

The identity substitution $1_{Γ}$ is defined as usual and $(i0)$ is
an abbreviation for $(1_{Γ},i=0)$. Also $b[u]$ is an abbreviation for
$b(1_{Γ},x=u)$.
\end{definition}

\begin{definition}
We define $tσ$ by induction on $t$

\begin{align*}
  Uσ &= U \\
  x(σ,f) &= xσ \\
  x(σ,y=u) &= xσ \\
  x(σ,x=u) &= u \\
  ((x:A)→ B)σ &= (y:Aσ) → B(σ,x=y) \\
  (λ x:A. b)σ &= λ y:Aσ. b(σ,x=y) \\
  (t.i)σ &= (t(σ,i=j)).j \\
  (f a)σ &= fσ  (aσ) \\
  (a,_i b)_T(σ,x=u)  &= (a,_i b)_Tσ  \\
  (a,_i b)_T(σ,f)  &= (a,_i p)_Tσ & i \# f \\
  (a,_i b)_T(σ,f)  &= a(σ,f-i) & f(i) = 0 \\
  (a,_i b)_T(σ,f)  &= (a(σ,f-i),_j b(σ,f-i))_{T(σ,f)} & f(i) = j
\end{align*}
\end{definition}

\subsection{Canonicity}

We have new normal forms:

\subsection{Swap and symmetric groups.}

\begin{lemma}\label{lem:param-var}
  $x.i.j = x.j.i$
\end{lemma}
\begin{proof}
  By $α$-equivalence of terms.
\end{proof}

Applying parametricity to many variables is useful and (by the above)
does not depend on the names of the variables. Hence we use the
notation $\pvar x k$ for $k$ applications of parametricity to the
variable $x$.

\begin{definition}
We define the following, well-typed term:
  $\sw {ij} T a q p r = (\CC i {\CC j a p T} {\CC j q r {T.i \_}} T) \param j \param i$

in the context.
\begin{align*}
  T &: U \\
  q &: T \param i a \\
  p &: T \param j a \\
  r &: T \param i \param j a p q \\
  \sw {ij} T a q p r & : T \param j \param i a q p \\
\end{align*}
\end{definition}

\begin{corollary}
  The relation $\pvar y 2 x$ is symmetric.
\end{corollary}

\begin{conjecture}
  Swapping is involutive.
\end{conjecture}

\subsection{Normal forms}

\begin{align*}
  n,m & ::= \pvar x k ~|~ n t \\
  t,u,A,B & ::= U ~|~ λx:A. t   ~|~  (x:A) → B
             ~|~ \CC i t u A ~|~ \CSig i A B
\end{align*}


\subsection{Semantics}

- Nominality; dependence on a finite number of colors.

Le modele correspond a pretendre connaitre toutes les expansions en hypercubes.

un type A sur i,j,... couleurs est

A : Set
A.i : A -> Set
A.j : A -> Set
A.i.j : (x:A) -> A.i x -> A.j x -> Set
...

mais A.j.i est donne par

A.j.i a p q = A.i.j a q p     (eq. 1)

Le swap est donc purement formel, et uniquement dans les types.

En effet, avoir a : A sur i,j,... couleurs est

a : A
a.i : A.i a
a.j : A.j a
a.i.j : A.i.j a a.i a.j
...

a.j.i est donne par a.i.j (pas de transformation)


\end{document}
